\documentclass{report}

% The ams packages are required to insert any mathematical symbols that you may require
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
% This package is used for embedding things like PDFs and JPEGs into your document
\usepackage{graphicx}
% This package is used for drawing pictures (such as trees)
\usepackage{tikz}
\usetikzlibrary{arrows,automata}
% These packages are used for adding pseudo-code to your document
\usepackage{algorithm2e}
\usepackage{algorithmic}
% This package is for if you require an appendix
\usepackage{appendix}

\usepackage{datetime}
\usepackage{fancyhdr}
\usepackage{semantic}
\usepackage[noload]{qtree}
\usepackage{comment}
\usepackage{pdflscape}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage{moreverb}
\usepackage{caption}
\usepackage{url}
\usepackage{layout}
\usepackage{subfig}
\usepackage{lineno}

\linenumbers

\usepackage[annatar]{tengwarscript}

\usetikzlibrary{shapes}
\usetikzlibrary{arrows}

\newcommand{\pr}{\mathbb{P}}
\newcommand{\tengt}{\mbox{\Ttinco}}
%\newcommand{\tengt}{\Gamma}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}

\begin{document}

\input{./title.tex}

\tableofcontents

\chapter{Introduction}

\input{./intro.tex}

\chapter{A Zoo of Stochastic Processes}\label{zoo}

\input{./processes.tex}

\chapter{Potential Modelling Techniques}\label{techniques}

\input{./techniques.tex}

\appendix
\chapter{Fun with Integrals}\label{fun}

\input{./appendix.tex}

\begin{thebibliography}{9}

\bibitem{mmpp1} Ihler,A.,Hutchins,J.,\& Smyth,P Learning to detect events with Markov-Modulated Poisson Processes.

ACM Transactions on Knowledge Discovery from Data, 1(3), 13. 2007 

\url{http://dl.acm.org/citation.cfm?doid=1297332.1297337}

\bibitem{mmpp2} R.D. Malmgren, D.B. Stouffer, A.E. Motter, and L.A.N. Amaral, A Poissonian explanation for heavy tails in e-mail communication, 

PNAS, 105(47):18153–18158, 2008 

\url{http://www.pnas.org/content/105/47/18153}

\bibitem{mmpp3} Scott, S. L. and Smyth, P. 2003. The Markov modulated Poisson process and Markov Poisson cascade with applications to web traffic data.

Bayesian Statistics 7, 671–680. 

\url{http://dl.acm.org/citation.cfm?doid=1297332.1297337}

\bibitem{doob96}
	Joseph L. Doob, The Development of Rigor in Mathematical Probability (1900-1950)

	The American Mathematical Monthly, Vol. 103, No. 7 (Aug. - Sep., 1996), pp. 586-595

	\url{http://www.jstor.org/stable/2974673}

\bibitem{mwmarkov}
	Weisstein, Eric W. ``Markov Chain." From MathWorld--A Wolfram Web Resource. 

	\url{http://mathworld.wolfram.com/MarkovChain.html}

\bibitem{mwgraph}
	Weisstein, Eric W. ``Graph." From MathWorld--A Wolfram Web Resource. 
	
	\url{http://mathworld.wolfram.com/Graph.html}

\bibitem{thinning}
	Lewis, P. A. W. and Shedler, G. S. (1979), Simulation of nonhomogeneous Poisson processes by thinning.

	Naval Research Logistics, 26: 403–413.

	\url{http://onlinelibrary.wiley.com/doi/10.1002/nav.3800260304/abstract}

\bibitem{hiddenmarkov}
	David Harte. \emph{HiddenMarkov} v1.7-0, a Hidden Markov Model library written in R and Fortran, 

	\url{http://cran.r-project.org/web/packages/HiddenMarkov/index.html}

\bibitem{matplotlib}
	Hunter, J. D. \emph{MatPlotLib}, a 2D graphics environment 

	\url{http://matplotlib.org/}

	IEEE Computer Soc, Vol. 9, No. 3 (2007), pp. 90-95

\bibitem{gimpdiff} The GIMP Development Team. \emph{The GIMP docs}, the documentation for the GNU Image Manipulation Program, 

\url{http://docs.gimp.org/en/gimp-concepts-layer-modes.html}

\bibitem{gimpthresh} The GIMP Development Team. \emph{The GIMP docs}, the documentation for the GNU Image Manipulation Program, 

\url{http://docs.gimp.org/en/gimp-tool-threshold.html}

\bibitem{rpy} Moriera W, Warnes GR, Gautier L. \emph{RPy} v2-2.3, a Python to R interface

\url{http://rpy.sourceforge.net/}

\bibitem{scipy} Jones E, Oliphant T, Peterson P \& others, \emph{SciPy}, open source scientific tools for Python, 2001-

\url{http://www.scipy.org}

\bibitem{baumwelch} Baum L.E, Petrie T, Soules G, \& Weiss N, A maximization technique occuring in the statistical analysis of probabilistic functions of Markov chains.

Annals of Mathematical Statistics 41(1), 164-171, 1970

\url{http://projecteuclid.org/DPubS?service=UI&version=1.0&verb=Display&handle=euclid.aoms/1177697196}

\bibitem{viterbi} Forney G.D, Jr., ``The viterbi algorithm,"

Proceedings of the IEEE , vol.61, no.3, pp.268,278, March 1973

\url{http://ieeexplore.ieee.org/stamp/stamp.jsp?tp=&arnumber=1450960&isnumber=31166}

\bibitem{bic} Schwarz, G. (1978) "Estimating the Dimension of a Model"

Annals of Statistics, 6, 461-464.

\url{http://projecteuclid.org/DPubS?service=UI&version=1.0&verb=Display&handle=euclid.aos/1176344136}

\bibitem{kmeans} Jones E, Oliphant T, Peterson P \& others, \emph{SciPy}'s k-means algorithm, \url{http://docs.scipy.org/doc/scipy/reference/cluster.vq.html}

\bibitem{mwlognormal} Weisstein, Eric W. "Log Normal Distribution." From MathWorld--A Wolfram Web Resource.

\url{http://mathworld.wolfram.com/LogNormalDistribution.html}

\bibitem{montecarlo} Caflisch E, Monte Carlo and quasi-Monte Carlo methods,

Acta Numerica vol. 7, Cambridge University Press, 1998, pp. 1–49.

\url{http://websrv.cs.fsu.edu/~mascagni/Caflisch_1998_Acta_Numerica.pdf}

\end{thebibliography}

\end{document}
